---
title: "Introduction To Empirical Bayes Notes"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
renv::restore()
r_packages <- c("ggplot2", "dplyr", "tidyr", "purrr")
renv::install(r_packages)

library(dplyr)
library(tidyr)
library(ggplot2)
library(purrr)
```

## Why empirical Bayes?

Empirical Bayesian methods are an approximation to more exact methods, but come with some controversy. However, EB methods are well suited for Data Science because

- Though EB is Inacurate with few observations, with large datasets performance can be comperable to traditional Baysiean methods
- EB provides shortcuts that allow easy computation at scale; Approximate in a fraction of the time of full Baysien Methos

We can use EB as a fast approximation technique, when we have many observations.

## The beta distribution

The _beta distribtion_ is a probability distribution with two paramters $\alpha$ (alpha) and $\beta$ (beta).  The distribution is constrained between 0 an 1.  In practice, it is good at representing a probabilty distribution of probabilities.

```{r, echo = FALSE}
size = 1e6
tibble::tibble(
  alpha = c(1, 3, 20, 50),
  beta  = c(2, 3, 20, 10),
) %>% dplyr::mutate(
  batting_average  = purrr::map2(alpha, beta, ~ rbeta(size, .x, .y)),
  Paramters = paste0("alpha = ", alpha, " beta = ", beta)
) %>% 
  tidyr::unnest(batting_average) %>%
  ggplot2::ggplot(ggplot2::aes(x = batting_average, color = Paramters)) + 
  ggplot2::geom_density() + 
  ggplot2::xlab("Batting Average") +
  ggplot2::ylab("Density of beta") + 
  theme_bw()
```

## The binomial distribution

The _binomial distribution_ models a count of successes out of the total.  To follow the baseball analogy, for 100 at-bats we would say the number of hits a player will get is distribted according to `Binomial(100, p)` where `p` is the probability that at-bat `AB` will be a hit `H`.  In the baseball case, the `p` the thing we want estimate is a batting average.

> Batting Average = Hits / At Bats

## Updating The Beta Distribution

We can build an existing belief (prior) regarding what a reasonable batting average should be.  The text suggests that belief is 0.27, with an alpha = 81, and beta of 219.  However, a batting average could reasonably range from 0.21 to 0.35 (see plot below).  So if we start with this existing belief of where any players true batting average _could be_, how do we update our prior beliefs based on new information as the player gets at-bats in.

```{r, echo = FALSE}
size = 1e6
tibble::tibble(
  alpha = c(81),
  beta  = c(219),
) %>% dplyr::mutate(
  batting_average  = purrr::map2(alpha, beta, ~ rbeta(size, .x, .y)),
  Paramters = paste0("alpha = ", alpha, " beta = ", beta)
) %>% 
  tidyr::unnest(batting_average) %>%
  ggplot2::ggplot(ggplot2::aes(x = batting_average, color = Paramters)) + 
  ggplot2::geom_density() + 
  ggplot2::xlab("Batting Average") +
  ggplot2::ylab("Density of beta") + 
  theme_bw()
```

This idea, `prior expectations + new information = new expectation` is a simplificiation of the Bayesian Philosopy.  In the case of updating our beta distribution, the update rule is very simple.

> Beta(alpha<sub>0</sub> + hits, beta<sub>0</sub> + misses)

So, if a player were to have a hit on their first at-bat, the new beta would be be Beta(82, 219).  When we plot both distributions we can see that the posterior distribution shifts slightly to the right.  The movement is very small, however, if we were to consider a batter who has hit 200 out of 600 at bats, the new posterier would be be Beta(282, 619), a much bigger shift, since we have collected more evidence for the update rule.

```{r}
size = 1e6
tibble::tibble(
  alpha = c(81, 82, 282),
  beta  = c(219, 219, 619),
) %>% dplyr::mutate(
  batting_average  = purrr::map2(alpha, beta, ~ rbeta(size, .x, .y)),
  Paramters = paste0("alpha = ", alpha, " beta = ", beta)
) %>% 
  tidyr::unnest(batting_average) %>%
  ggplot2::ggplot(ggplot2::aes(x = batting_average, color = Paramters)) + 
  ggplot2::geom_density() + 
  ggplot2::xlab("Batting Average") +
  ggplot2::ylab("Density of beta") + 
  theme_bw()
```

## Pos
